Personalization and probabilities: impersonal propensities in online grocery shopping

### 2017-10-02 log.rmd
fixing up lost manuscript bits and references

>>>>

## Mon 02 Oct 2017 12:14:12 BST
- facing resubmission build. Bound to be missing references. I have them all somewhere.  
- using render_tufte.r
- drats - problem with the work I did on Friday -- some kind of crash (battery ran out?) has lost work I did later on Friday morning. Need to redo that now. The TBAs are still there for instance. 
- have got most of the references in; need to get Christian_2012 in there. 
- TODO: go abstract and intro again;
    - go through conclusion again;
    - redo TBAs
    - check whether the extra Amoore point about homogeneous calculative space is there

### 2017-10-02 reconstruction.rmd
fixing up lost manuscript bits and references

Much big data practice converts an existing set of practices by adding a model or predictive elements based on the collection and analysis of data. This paper recounts the conversion of  an online grocery ordering system  from an older demographically-based recommendation system to a 'personal relevance' model. The account is empirical in several respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how personalized recommendations were created for online grocery orders.  The paper  draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping and lists. It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] to identify and map functional statements and practices configuring the knowledge and order generated by such systems over time. It conducts several small-scale code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some prototypical elements of the system in question.  Finally,  it  draws  on the philosophy of science to frame an engagement with grocery recommendations in terms of probabilities and their worldly enactments.

The main argument of the paper concerns how  personalization occurs in big data 'conversion events.' The term 'conversion event' has a dual sense. In recommender system design and web ecommerce, it refers occasions when a customer, user or site visitor clicks or purchases. I also use it to   refer to the common before-after narrative forms that proponents, developers and advocates of recommender systems adopt in describing, promoting, explaining or otherwise talking about big data. In recent research and opinion on data practice, personalization has been seen as the linchpin of predictive modelling.  Big data discourse in its promissory mode  attributes potency to personalization: 'most important, using big data we hope to identify specific individuals rather than groups' [@Mayer-Schonberger_2013]. Conversely, Joseph Turow and co-authors conclude in their critical account of the transformation of retail space: 'through it all, knowingly and not, and away from the spotlights of fierce social debate, retailers are encouraging daily routines that accept data-driven personalization as a centrifugal public' [@Turow_2015, 476]. Analysis of internet filter bubbles and the growth of predictive platforms in industry and government  [@Pariser_2011; @ONeil_2016a] also centre on personalization. In this paper, I agree that a transformation or conversion event is occurring, but one which also, even in the midst of much ongoing personalization,  opens the possibility of conceiving of social ordering processes afresh. The average everydayness and familiarity of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- social media for instance -- do not.

'Data-driven personalization' relies on predictions, themselves dependent on data practices ranging across acquisition, storage, transformation, exchange, integration, modelling and experimentation.  The reconstruction of a recommender system and its  personalized predictions I undertake here focuses on the notion of probability and its role in modelling and experimentation.  Although probabilistic calculations of events have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on an operational function in digital platforms associated with the web and social media.  They have become part of infrastructures, transactions, and event-structures. In order this _operational_ reality into account, we might understand probabilities in quasi-physicalist terms  defined by the philosopher of science Karl Popper as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. This operational-physicalist or sociotechnical notion of probability shifts our understanding of data analytics, especially in association with platforms [@Gillespie_2010], away from personalization.

Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Marcus_2014; @Bogost_2012], the paper allows its own reconstruction to be provisionally shaped by an experiential encounter with the object it describes. At times, this mode of empirical philosophy may seem fixated on the technical minutiae of algorithms and recommender systems. With some forbearance on the part of readers, however, the argument of the paper should resonate sociologically. It concerns the relationality of social life, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities play a pervasive role.

@@

Against the background of the sheer number of commodities and their distribution of prices,  Patel's presentation presents two opportunities. First, by recounting, contextualising and commenting on the main steps in making the 'Have your forgotten?' list, we might follow some of the predictive sense-making done by data scientists and customer analytics teams working with transactional data in a typical commercial setting. Patel mentioned many of these steps only fleetingly in the presentation, for they are largely taken for granted as part of predictive analytic practice. Second, Patel focused on the renovation and updating of long-standing data-mining practice via a much more explicitly 'big data' and 'machine learning'- oriented implementation. Her presentation concerns what  a typical big data conversion event  in which a long-standing recommender system was replaced by a 'big data'-style system delivering 'personally relevant' recommendations.   What stands out from the presentation is the normality of the recommender system: it is part of a long-standing and ongoing transformation of grocery shopping. From a sociological standpoint, the interest lies less in specific technical innovations and more on how the ordering work done by recommender systems relates to the role of prediction in social order more generally.

@@

The main narrative of Patel's conversion narrative concerned the shift from a well-established loyalty-card based data-mining model developed in the 1990s to a predictive, probabilistic, 'personal relevance' model that would  append to the shopping list in almost-realtime. While the changes Patel described are configured in Tesco-specific ways by DunnHumby, they are also broadly typical. Tesco is the largest supermarket chain in UK.  Its  customer loyalty and targeted marketing programme known as 'Tesco Clubcard' started in 1991. DunnHumby -- founded by operations researchers Edwina Dunn and Clive Humby -- is said to have convinced the CEO of Tesco sometime in 1991 that a loyalty card program could change the supermarket chain's relationship to its customers.  Clive Humby's academic publications are hard to track down.   An early paper given at the _Conference of Young Operational Researchers_ in Nottingham in 1984 suggests the direction that he, the company Dunn and Humby formed (DunnHumby) and later Tesco would take in constructing lists [@OKeefe_1984]. The abstract for Humby's presentation pre-figures an ongoing trajectory for data mining techniques aimed at eliciting detailed information on individual customer references.

@@

In  [@Humby_1989], Humby argues for the need to add lifestage data to neighbourhood data in marketing research. While Tesco  succeeded in data-mining its customers using demographic segmentation,  and perhaps became the UK's biggest supermarket with the help of data-mining in the 1990s, the shopping environment  in 2017 is markedly different [@Turow_2017]. It is no longer organised around campaigns involving special offers or redemption of points for demographically segmented loyal customers (DunnHumby made heavy use of UK Census data). It can no longer rely on placement of goods in carefully chosen locations in stores. It needs, or at least might want to, up-sell and cross-sell to customers who only sometimes visit the supermarket itself.

How could we characterise the shift from ClubCard data mining to online grocery markets?  If 'Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet' [@Silverwood-Cope_2014], then has Tesco itself undergone some kind of conversion? Patel described her work at DunnHumby as converting the recommender system from a 'rules-based list' to a 'relevancy model.' The relevancy model affected the construction of the 'Have You Forgotten' list.   This very localised intervention is typical of broader re-organisation of prediction. The shift in models results in a more probabilistic structuring of lists.  As is so often the case in big data conversion narratives, the triviality of the 'Have you forgotten?' recommendations provides only peripheral signs of the complex predictive infrastructure underpinning them.

@@

Viewed from the standpoint of probabilities, predictive systems are highly crafted arrangements for the calculation of conditional probabilities. For instance, the first element of the new recommender system consisted in a change of the underpinning algorithms and model. Patel described  move away from 'a rules-based system'. It is likely that what Patel describes as  the 'rules-based system' refers  to the extremely well-known association rules or `apriori` algorithm  learning technique, developed  by computer scientists Rakesh Agrawal and Ramakrishnan Srikanti working at IBM Research Alameda in the early 1990s [@Agrawal_1994]. A now-classic approach to 'market basket analysis,' it was listed as one of the top ten data mining algorithms in a survey conducted amongst data miners [@Wu_2008] and usually attracts a chapter in data-mining and machine learning textbooks (e.g. [@Hastie_2008]). The interest of `apriori` for our purposes is that it begins to address the problem of understanding large numbers of shopping transactions as a matter of conditional probability.

@@

The notion of conditional probability at work in `aprior` is relatively simple, and assumes that the frequency of transactions provides the best guide to what shoppers are likely to buy. The `apriori` algorithm finds sets of items that commonly occur together in transactions. In this sense, is still oriented by a notion of absolute probability, inflected by some elements of conditional probability.   Commonly occurring sets are expressed as 'association rules.' For instance, applied to a dataset of generic, unbranded grocery purchases, the `apriori` algorithm counts frequencies of purchase in the overall set of all items purchased in a supermarket (the `Groceries` dataset  was acquired from a 'local German supermarket' [@Hahsler_2006a]). Figure \@ref(fig:arules-ex) shows how often the most common items appear. `Whole milk` appears most frequently.

@@

The possibility of adjusting the recommendations for each customer every week depends on an infrastructure capable of collecting data, and assimilating that data in a predictive model.  'Personal relevance' depends on a matrix of probabilities of associations between people and things that shifts in time.  `Hadoop` and its legion of 'big data' variants (`mahout`, `spark`, `hive`, `pig`, `yarn`) operationalise repetition at an infrastructural rather than analytic scale. Patel's quick gloss of the infrastructural deployment of DunnHumby's relevancy model is  primary to the conversion event: the logistic regression model at the heart of the recommender system is no longer an analytical device but an operational one because it seeks to revise recommendations as situations change. [TBA: this paragraph is promising but doesn't deliver yet]

@@

Like the accounts of laboratory practice provided by science studies [@Latour_1097; @Rheinberger, @Knorr-Cetina; @Lynch_1994], Popper understands experiments as stabilizing the conditions so that probabilities become less variable. Experiments play an important role in conversion events, or in the process of bring new predictive models to bear on shopping practice. A crucial consideration for  a recommender system designer will be how to experimentalize the systems recommendations so that they are not disturbed by other propensities. This is an inordinately difficult challenge, since, as we have seen, grocery shopping takes place at the intersection of normative and actual social orders, individual and group belongings, and logistically global economic processes.

|Test A|Test B|
|------|------|
|Control A|Control B|
---------------------

@@

Random allocation of customers to control and test groups occurs without taking individual propensities into account.  A/B testing  seeks to statistically validate effects -- the *uplift* -- of the model on  conversion rates by directly measuring the effects of the model on what people do.  The uplift refers to conversion events associated with the same groups of people. Effectively an experiment in creating micrologically different worlds, the randomised control trial  sets up a control mechanism that connects  the predictive model (the logistic regression), and the conversion event more broadly. With this experimentalisation, the conversion event narrative lacks grounding.[@TBA: not happy with this ending]

@@

I suggest that we might understand what precedes and exceeds personalization as probabiliization.  If, as Popper argues, probabilities are being realized in the same way that 'fields' were realized in modern physics, then we should see the compute  platforms, databases, software libraries, web interfaces and global supply chain logistics as part of this conversion event. When we observe what is practically done to construct a personal relevance model and apply it to a shopping list, we see conditional probabilities assembled in predictive models, the rhythms of  platform and infrastructural configurations in gathering and processing data,  and the open-ended construction of features in data. In various ways, they shift and complicate the  associations between people and things, people and people, and things with things. If social order is made of associations, if to be social is to associate, then big data conversion events operationalize association in matrices of propensity.

>>>>

Big data practice often converts an existing state of affairs by adding a model or predictive elements based on the collection and analysis of data. This paper recounts the conversion of  an online grocery ordering system  from an older demographically-based recommendation system to a 'personal relevance' model. The main argument of the paper concerns how  personalization occurs in big data 'conversion events.' The term 'conversion event' has a dual sense. In recommender system design and web ecommerce, it refers to those occasions when a customer, user or site visitor clicks a specific hyperlink or purchases something. I also use it to refer to the common before-after narrative forms that proponents, developers and advocates of recommender systems adopt in describing, promoting, explaining or otherwise talking about big data, analytics or predictive modelling practices. In recent research and opinion on recommendations, personalization has been seen as the goal of predictive modelling.  Big data discourse in its promissory mode attributes potency to personalization: 'most important, using big data we hope to identify specific individuals rather than groups' [@Mayer-Schonberger_2013]. Conversely, current debates about the problems of big data emphasize the need to protect people from personalization.  Joseph Turow and co-authors conclude in their critical account of the transformation of retail space by business analytics: 'through it all, knowingly and not, and away from the spotlights of fierce social debate, retailers are encouraging daily routines that accept data-driven personalization as a centrifugal public' [@Turow_2015, 476]. Analysis of internet filter bubbles and the growth of predictive platforms in industry and government  [@Pariser_2011; @ONeil_2016a] also centre on personalization. In this paper, I agree that a transformation or conversion event is occurring, but one which also, even in the midst of much ongoing personalization,  opens the possibility of conceiving of social ordering processes afresh. Like Louise Amoore and Viola Piotukh's call for a new epistemology of population  [@Amoore_2015, 360], I will argue that contemporary recommender systems afford us the opportunity to re-conceptualise long-standing assumptions about social order and structure. The average everydayness and familiarity of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- social media for instance -- do not.

'Data-driven personalization' relies on predictions, themselves dependent on data practices ranging across acquisition, storage, transformation, exchange, integration, modelling and experimentation.  The reconstruction of a recommender system and its  personalized predictions that I undertake here centres on the notion of probability and its role in modelling and experimentation.  Although probabilistic calculations of events have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on an operational function in digital platforms associated with the web and social media.  They have become part of infrastructures, transactions, and event-structures. In order to take this _operational_ reality into account, we might understand probabilities in quasi-physicalist terms  defined by the philosopher of science Karl Popper as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. This operational-physicalist or sociotechnical notion of probability shifts our understanding of data analytics, especially in association with platforms [@Gillespie_2010], away from personalization. Finally,  it  draws  on the philosophy of science to frame an engagement with grocery recommendations in terms of probabilities and their worldly enactments.

The account is empirical in several respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how personalized recommendations were created for online grocery orders.  The paper  draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping and lists. It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] to identify and map functional statements and practices configuring the knowledge and order generated by such systems over time. It conducts several small-scale code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some prototypical elements of the system in question.  Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Marcus_2014; @Bogost_2012], the paper is  shaped by practical encounter with the 'object' -- code, data, web platforms --  it describes. At times, this mode of empirical philosophy may seem overly fixated on the technical minutiae of algorithms and recommender systems and inattentive to lived experience or practices. With some forbearance on the part of readers, however, the argument of the paper should resonate sociologically. It concerns the relationality of social life, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities are widespread.

@@

Against the background of the sheer number of commodities and their distribution of prices, Patel's presentation presents two opportunities. First, by recounting, contextualising and commenting on the main steps in making the 'Have your forgotten?' list, we might follow some of the predictive sense-making done by data scientists and customer analytics teams working with transactional data in a typical commercial setting. Patel mentioned many of these steps only fleetingly in the presentation, for they are largely taken for granted as part of predictive analytic practice. Second, Patel focused on the renovation and updating of long-standing data-mining practice via a much more explicitly 'big data' and 'machine learning'- oriented implementation. Her presentation concerns a typical big data conversion event  in which a long-standing recommender system was replaced by a 'big data'-style system delivering 'personally relevant' recommendations.   What stands out from the presentation is the normality of the recommender system: it is part of a long-standing and ongoing transformation of grocery shopping. From a sociological standpoint, the interest lies less in specific technical innovations and more on how the ordering work done by recommender systems relates to the role of prediction in social order more generally.

@@

The main narrative of Patel's conversion narrative concerned the shift from a well-established loyalty-card based data-mining model developed in the 1990s to a predictive, probabilistic, 'personal relevance' model that would  append items to the shopping list in almost-realtime. While the changes Patel described are configured in Tesco-specific ways by DunnHumby, they are also broadly typical of a big data conversion event. Analytics service providers such as DunnHumby attempt to convert their customers by stories of conversion.[^65.1] Patel's presentation was part of this effort. Tesco is the largest supermarket chain in UK and a notable success story for DunnHumby.  Tesco's  customer loyalty and targeted marketing programme known as 'Tesco Clubcard' started in 1991. DunnHumby -- founded by operations researchers Edwina Dunn and Clive Humby -- is said to have convinced the CEO of Tesco sometime in 1991 that a loyalty card program could change the supermarket chain's relationship to its customers.[^65.2]  Clive Humby's academic publications are hard to track down.   An early paper given at the _Conference of Young Operational Researchers_ in Nottingham in 1984 suggests the direction that he, the company Dunn and Humby formed (DunnHumby) and later Tesco would take in constructing lists [@OKeefe_1984]. The abstract for Humby's presentation pre-figures an ongoing trajectory for data mining techniques aimed at eliciting detailed information on individual customer references.

[^65.1] The web pages of the big data analytic companies providing products to various sectors of industry, government and media are stocked with images, graphics and statements centred on this conversion experience. [@Amoore_2015] analyses some of these.

[^65.2]: Tesco changed its relations to customers in 1984 when they launched a very online grocery shopping system [@Winterman_2013].

@@

In  [@Humby_1989], Humby highlights the need to add lifestage data to neighbourhood data in marketing research. While Tesco  succeeded in data-mining its customers using demographic segmentation,  and perhaps became the UK's biggest supermarket with the help of data-mining in the 1990s, the shopping environment  in 2017 is markedly different [@Turow_2017]. It is no longer organised around campaigns involving special offers or redemption of points for demographically segmented loyal customers (DunnHumby made heavy use of UK Census data). It can no longer rely on placement of goods in carefully chosen locations in stores. It needs, or at least might want to, up-sell and cross-sell to customers who only sometimes visit the supermarket itself.

How could we characterise the shift from ClubCard data mining to online grocery markets?  If 'Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet' [@Silverwood-Cope_2014], then has Tesco itself undergone some kind of conversion? Patel described her work at DunnHumby as converting the recommender system from a 'rules-based list' to a 'relevance model.' The relevance model affected the construction of the 'Have You Forgotten' list.   This very localised intervention is typical of broader re-organisation of prediction. The shift in models results in a more probabilistic structuring of lists.  As is so often the case in big data conversion narratives, the triviality of the 'Have you forgotten?' recommendations provides only peripheral signs of the complex predictive infrastructure underpinning them.

@@

Viewed from the standpoint of probabilities, predictive systems are highly crafted arrangements for the calculation of conditional probabilities. For instance, the first element of the new recommender system consisted in a change of the underpinning algorithms and model. Patel described  move away from 'a rules-based system'. It is likely that what Patel describes as  the 'rules-based system' refers  to the extremely well-known association rules or `apriori` algorithm  learning technique, developed  by computer scientists Rakesh Agrawal and Ramakrishnan Srikanti working at IBM Research Alameda in the early 1990s [@Agrawal_1994]. A now-classic approach to 'market basket analysis,' it was listed as one of the top ten data mining algorithms in a survey conducted amongst data miners [@Wu_2008] and usually attracts a chapter in data-mining and machine learning textbooks (e.g. [@Hastie_2009]). The interest of `apriori` for our purposes is that it begins to address the problem of understanding large numbers of shopping transactions as a matter of conditional probability.

@@

The notion of conditional probability at work in `aprior` is relatively simple, and assumes that the frequency of transactions provides the best guide to what shoppers are likely to buy. The `apriori` algorithm finds sets of items that commonly occur together in transactions. In this sense, is still oriented by a notion of absolute probability, inflected by some elements of conditional probability.   Commonly occurring sets are expressed as 'association rules.' For instance, applied to a dataset of generic, unbranded grocery purchases, the `apriori` algorithm counts frequencies of purchase in the overall set of all items purchased in a supermarket (the `Groceries` dataset  was acquired from a 'local German supermarket' [@Hahsler_2006]). Figure \@ref(fig:arules-ex) shows how often the most common items appear. `Whole milk` appears most frequently.

@@

The possibility of adjusting the recommendations for each customer every week depends on an infrastructure capable of collecting data, and assimilating that data in a predictive model.  'Personal relevance' depends on a matrix of probabilities of associations between people and things that shifts in time.  `Hadoop` and its legion of 'big data' variants (`mahout`, `spark`, `hive`, `pig`, `yarn`) operationalise repetition at an infrastructural rather than analytic scale. Patel's quick gloss of the infrastructural deployment of DunnHumby's relevance model is  primary to the conversion event: the logistic regression model at the heart of the recommender system is no longer an analytical device but an operational one because it seeks to revise recommendations as situations change.

@@

Like the accounts of laboratory practice provided by science studies [@Latour_1986;  @Lynch_1993], Popper understands experiments as stabilizing the conditions so that probabilities become less variable. Experiments play an important role in conversion events, or in the process of bring new predictive models to bear on shopping practice. A crucial consideration for  a recommender system designer will be how to experimentalize the systems recommendations so that they are not disturbed by other propensities. This is an inordinately difficult challenge, since, as we have seen, grocery shopping takes place at the intersection of normative and actual social orders, individual and group belongings, and logistically global economic processes.

|Test A|Test B|
|------|------|
|Control A|Control B|
|---------------------

@@

Random allocation of customers to control and test groups occurs without taking individual propensities into account.  A/B testing  seeks to statistically validate effects -- the *uplift* -- of the model on  conversion rates by directly measuring the effects of the model on what people do.  The uplift refers to conversion events associated with the same groups of people. Effectively an experiment in creating micrologically different worlds, the randomised control trial  sets up a control mechanism that connects  the predictive model (the logistic regression), and the conversion event more broadly. With this experimentalisation, the conversion event narrative lacks grounding.[TBA: not happy with this ending]

@@

I suggest that we might understand what precedes and exceeds personalization as probabilization.  If, as Popper argues, a propensity-based account of probabilities 'amounts to generalizing and extending the idea of forces again' [@Popper_1990,14], then we should see the compute  platforms, databases, software libraries, web interfaces and global supply chain logistics as part of this conversion event, as a tendency in the process of realization. When we track what is practically done to construct a personal relevance model and apply it to a shopping list, we see conditional probabilities assembled in predictive models, the rhythms of  platform and infrastructural configurations in gathering and processing data,  and the open-ended construction of features in data. In various ways, they shift and complicate the  associations between people and things, people and people, and things with things. The framing of this conversion as a 'flat difference of degree, such that it appears as though everything is calculable' [@Amoore_2015, 361] risks misunderstanding the different propensities tending to realization.  If social order is made of propensities to associate, if to be social is a propensity to associate, then big data conversion events operationalize association in matrices of propensity.
